---
title: "N741 Homework 8 - Answer Key"
author: "Melinda K. Higgins, PhD."
date: "April 13, 2017"
output:
  pdf_document: default
  html_document: default
  github_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Homework 8 - DUE April 12, 2017 at 5pm

Please submit Homework 8 as a PDF to CANVAS no later than 5pm EST on April 12, 2017.

## Wisconsin Breast Cancer Data (Original)

For Homework 8 you will be working with the "Original" Wisconsin Breast Cancer dataset from the UCI Machine Learning Repository; see UCI dataset [http://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Original%29](http://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Original%29).

The raw data files can be downloaded from the associated Data Folder at [http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/](http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/). In this homework you will be working with the "breast-cancer-wisconsin.data" dataset, which is a CSV comma delimited file with NO column names in the 1st row. The datafile description and associated column file names are in the "breast-cancer-wisconsin.names" which is a simple text file [http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names](http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names). In this text file, as you read through it and scroll down, you'll see the following:

```
7. Attribute Information: (class attribute has been moved to last column)

   #  Attribute                     Domain
   -- -----------------------------------------
   1. Sample code number            id number
   2. Clump Thickness               1 - 10
   3. Uniformity of Cell Size       1 - 10
   4. Uniformity of Cell Shape      1 - 10
   5. Marginal Adhesion             1 - 10
   6. Single Epithelial Cell Size   1 - 10
   7. Bare Nuclei                   1 - 10
   8. Bland Chromatin               1 - 10
   9. Normal Nucleoli               1 - 10
  10. Mitoses                       1 - 10
  11. Class:                        (2 for benign, 4 for malignant)
```

So, the final datafile will have 11 columns. The dataset itself is a compilation of multiple groups of clinical cases also detailed in the breast-cancer-wisconsin.names" file [http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names](http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names).

The combined dataset has 699 cases (rows). However, 16 cases were missing values for the "Bare Nuclei" measurement. The R code below, processes the data, applies the names, and removes the cases with missing values. So, the final dataset created below `bcdat` will have 683 cases and 11 variables.

```{r}
# from tidyverse - use readr
# to read in the comma delimited dataset
library(readr)

# raw data does not have column names
bcdat <- read_csv("breast-cancer-wisconsin.data",
                  col_names=FALSE)

# add variable names
names(bcdat) <- c("idnum","clumpthickness","uniformcellsize",
                  "uniformcellshape","marginaladhesion",
                  "singlecellsize","barenuclei","blandchromatin",
                  "normalnucleoli","mitoses","class")

# note in column 7 "Bare Nucleoli" there are
# question marks "?" that need to be set to missing NA
library(dplyr)
bcdat <- bcdat %>%
  mutate(barenucfix = ifelse(barenuclei=="?",NA,
                              as.numeric(barenuclei)))

# keep the main 11 variables
bcdat <- bcdat %>%
  select(idnum,clumpthickness,uniformcellsize,uniformcellshape,
         marginaladhesion,singlecellsize,barenucfix,blandchromatin,  
         normalnucleoli,mitoses,class)

# keep only complete cases, n=683
bcdat <- na.omit(bcdat)
```

## Principal Components Analysis

For this Homework, please refer back to the code and exercises that Dr. Hertzberg presented during lesson 10 - specifically review towards the end of "Lesson10Part3.Rmd" see [https://github.com/vhertzb/Lesson10/blob/master/Lesson10Part3.Rmd](https://github.com/vhertzb/Lesson10/blob/master/Lesson10Part3.Rmd). During this exercise, Dr. Hertzberg introduced you to the `prcomp` procedure for performing principal components analysis. `prcomp` is part of the built-in `stats` package with base R. To learn morn type `help(prcomp)`.

In Dr. Hertzberg's example, she provided code for:

* performing the principal components analysis (pca)
* using the pca output to make a plot of the variances for each principal component (pc)
* computing the PVE (percent variance explained) and plotting the PVE
* and plotting the principal component "scores" of the cases (e.g. the "scores" plot)

I will layout the code below for running the PCA for the dataset as a whole, which will include also making a "loadings" plot for the variable "coefficients" or "loading weights" for each PC - these "loading plots" give us additional insight into (a) how the variables cluster or relate/correlate with each other or not and (b) where they fall in terms of relevance for each PC in the plot. For this dataset, we can easily get away with keeping only 2 PCs and making simplier 2D scatterplots for both the "loading plot" and "scores plot".

Use the code steps below to help you complete this homework 8 assignment.

## 1. Perform the PCA

```{r}
# use only columns 2 through 10
# you do not need the idnum, nor the class variables
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Make plots of the variance and PVE 

### Plot of the Variances of Each PC

```{r}
plot(pr.out)
```

### Plot of the PVE and Cumulative PVE of each PC

```{r}
# plots of the PVE percent variance explained
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Make a "loadings plot" of the variables

```{r}
# loadings are in the "rotation" part of the 
# pr.out list object. "rotation" is a matrix
# with a row for each variable and a column for
# each PC.
pr.out$rotation

# choose the 1st and 2nd columns for the 1st 2 PCs
# and plot these loading weights for the 9
# variables. I tweaked the limits some
# feel free to change these as needed
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")

# add xpd=FALSE to prevent lines drawn outside plot area
par(xpd=FALSE)

# add red dashed lines for the axes at y=0 and x=0
abline(h=0, col="red")
abline(v=0, col="red")

# overlay the variable names on this loading plot
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Scores Plot on 1st 2 PCs

```{r}
# scores plot - use x from the pr.out list object
# plot scores on 1st 2 PCs, columns 1 and 2 of x
# color the points by the "class" variable for
# benign (class=2) or malignant (class=4)
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdat$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2",
     sub = "Blue=Benign (class=2) and Red=Malignant (class=4)")
```

## Homework 8 Tasks

1. Rerun the PCA (steps 1-4 above) for (A) just the Benign cases and for just the (B) Malignant Cases. The code below, sets up these data subsets for you.

```
# Benign cases
bcdatBenign <- bcdat %>%
  filter(class == 2)

# Malignant cases
bcdatMalignant <- bcdat %>%
  filter(class == 4)
```

_HINT: simply rename the new subsets and run the code steps above._ 

```
# redo for benign ==============
bcdat <- bcdatBenign
# run steps above

# redo for malignant ==================
bcdat <- bcdatMalignant
# run steps above
```

2. In the overall dataset, when looking at the loadings plot, which variables cluster together? which variables do not lie with that cluster?

3. How do the variable clusters seen in the loading plots for the Benign data subset and Malignant subset differ? and how are they similar if at all?

4. Is using 2 principal components reasonable for summarizing the variability seen in this Breast Cancer dataset with 9 measurements? Explain your reasoning for (a) the overall dataset, (b) the Benign subset and (c) the Malignant subset

5. While PCA is an unsupervised data analysis method (i.e. no "target" class information is used in the analysis), do you think the 2 PCs extracted do a good job of helping to distinguish Benign cases from Malignant cases (i.e. look back at the overall dataset Scores Plot). Explain your rationale.

6. Please save your RMD to a Github repository. Submit the PDF report for Homework 8 to CANVAS and include a link to your Homework 8 Github Repository.

----- 

## **ANSWER KEY**
## Task 1 (A): Rerun the PCA (steps 1-4 above) for (A) just the Benign cases 
**HINT** The easiest way to do this is after creating the subset, reassign the Bening subset back into the data object we used before `bcdat` and then simply rerun all the code again for steps 1-4. This avoid typos and forgetting to make updates in every code step.

```{r}
# save a copy before overwriting below
bcdatcopy <- bcdat

# Benign cases
bcdatBenign <- bcdat %>%
  filter(class == 2)

# redo for benign ==============
bcdat <- bcdatBenign
# run steps 1-4
```

## 1. Perform the PCA - BENIGN Data Subset

```{r}
# use only columns 2 through 10
# you do not need the idnum, nor the class variables
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Make plots of the variance and PVE - BENIGN Data Subset

### Plot of the Variances of Each PC

```{r}
plot(pr.out)
```

### Plot of the PVE and Cumulative PVE of each PC

```{r}
# plots of the PVE percent variance explained
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Make a "loadings plot" of the variables - BENIGN Data Subset

```{r}
# loadings are in the "rotation" part of the 
# pr.out list object. "rotation" is a matrix
# with a row for each variable and a column for
# each PC.
pr.out$rotation

# choose the 1st and 2nd columns for the 1st 2 PCs
# and plot these loading weights for the 9
# variables. I tweaked the limits some
# feel free to change these as needed
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")

# add xpd=FALSE to prevent lines drawn outside plot area
par(xpd=FALSE)

# add red dashed lines for the axes at y=0 and x=0
abline(h=0, col="red")
abline(v=0, col="red")

# overlay the variable names on this loading plot
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Scores Plot on 1st 2 PCs - BENIGN Data Subset

```{r}
# scores plot - use x from the pr.out list object
# plot scores on 1st 2 PCs, columns 1 and 2 of x
# color the points by the "class" variable for
# benign (class=2) or malignant (class=4)
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdat$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2",
     sub = "Blue=Benign (class=2) and Red=Malignant (class=4)")
```

## Task 1 (B): Rerun PCA (steps 1-4) for just the (B) Malignant Cases. 

```{r}
# reset bcdat
bcdat <- bcdatcopy

# Malignant cases
bcdatMalignant <- bcdat %>%
  filter(class == 4)

# redo for malignant ==================
bcdat <- bcdatMalignant
# run steps 1-4
```

## 1. Perform the PCA - MALIGNANT Data Subset

```{r}
# use only columns 2 through 10
# you do not need the idnum, nor the class variables
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Make plots of the variance and PVE - MALIGNANT Data Subset 

### Plot of the Variances of Each PC

```{r}
plot(pr.out)
```

### Plot of the PVE and Cumulative PVE of each PC

```{r}
# plots of the PVE percent variance explained
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Make a "loadings plot" of the variables - MALIGNANT Data Subset

```{r}
# loadings are in the "rotation" part of the 
# pr.out list object. "rotation" is a matrix
# with a row for each variable and a column for
# each PC.
pr.out$rotation

# choose the 1st and 2nd columns for the 1st 2 PCs
# and plot these loading weights for the 9
# variables. I tweaked the limits some
# feel free to change these as needed
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")

# add xpd=FALSE to prevent lines drawn outside plot area
par(xpd=FALSE)

# add red dashed lines for the axes at y=0 and x=0
abline(h=0, col="red")
abline(v=0, col="red")

# overlay the variable names on this loading plot
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Scores Plot on 1st 2 PCs - MALIGNANT Data Subset

```{r}
# scores plot - use x from the pr.out list object
# plot scores on 1st 2 PCs, columns 1 and 2 of x
# color the points by the "class" variable for
# benign (class=2) or malignant (class=4)
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdat$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2",
     sub = "Blue=Benign (class=2) and Red=Malignant (class=4)")
```

-----

## ANSWER KEY - Remaining questions

2. In the overall dataset, when looking at the loadings plot, which variables cluster together? which variables do not lie with that cluster?

**ANSWER KEY**

* In the original full dataset, all of the variables cluster together except for mitoses. 

-----

3. How do the variable clusters seen in the loading plots for the Benign data subset and Malignant subset differ? and how are they similar if at all?

**ANSWER KEY**

* A very similar cluster pattern is also seen for the Benign subset where most the the variables cluster together along PC1 with `mitoses` off by itself loading higher on PC2.
* However, for the Malignant data subset, the clusters are not as pronounced and show more variability. For example `clumpthickness` and `barenucfix` are somewhat separated from the rest of the variables and plot closer to PC2 and have higher loading weights for PC2 than for PC1. The other variables `uniformcellsize`, `uniformcellshape`, `singlecellsize`, `normalnucleoli`, `blandchromatin`, `marginaladhesion` all plot pretty close to PC1. `marginaladhesion` actually has a higher loading weight on PC2 but does ok on PC1 also - so this one is sort of loaded on both PC1 and 2. `mitoses` is somewhat closer to these variables, but the loading weights for `mitoses` were <0.3 on both PC1 and PC2, so `mitoses` as a variable does not load well on either PC1 nor PC2 for this malignant subset.

-----

4. Is using 2 principal components reasonable for summarizing the variability seen in this Breast Cancer dataset with 9 measurements? Explain your reasoning for (a) the overall dataset, (b) the Benign subset and (c) the Malignant subset

**ANSWER KEY**

* For the overall dataset, the 1st 2 PCs explain 74.172% of the variance in the dataset which is pretty good. There is also a significant drop off in PVE from PC1 to PC2 indicating that 1 PC might be enough to explain the majority of the variability in the dataset and 2 is even better.
* For the Benign subset, the 1st 2 PCs explain 47.58% of the variance in the dataset which is ok - it's not as good as the overall dataset though. There is still a sharp drop off in PVE from PC1 to PC2 indicating that 1 PC might be enough to explain the majority of the variability in the dataset and 2 is even better.
* For the Malignant subset, the 1st 2 PCs explain 46.11% of the variance in the dataset which is ok. There is a drop off in PVE from PC1 to PC2 indicating that the 1st PC captures most of the variance, but there is a steady decline in PVE from PC2 to PC9. 2 PCs do an ok job here; 3 might be better.

-----

5. While PCA is an unsupervised data analysis method (i.e. no "target" class information is used in the analysis), do you think the 2 PCs extracted do a good job of helping to distinguish Benign cases from Malignant cases (i.e. look back at the overall dataset Scores Plot). Explain your rationale.

**ANSWER KEY** When looking back at the Scres Plot for the Overall dataset with RED points for the Malignant cases and BLUE for the Benign cases, there is good visual separation between the 2 classes. There is some overlap along PC1 so it is possible that other classifiers might do a better job, but it is actually not that bad.

**EXTRA TIDBIT** You could even use the scores along PC1 to classify cases as Benign or Malignant - for example if we set cases with a PC1 score < 0.45 to Malignant which as you see in the plot below is where the density curves for the PC1 scores by class cross each other.

```{r}
# get the original overall dataset
bcdat <- bcdatcopy

# get the pc results and scores
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)

# add pc1 scores back to dataset
bcdat$pc1scores <- pr.out$x[,1]
bcdat$pc1malig <- bcdat$pc1scores < 0.45

# plot PC1 scores by original class
library(ggplot2)
ggplot(bcdat, 
       aes(x=pc1scores, color=as.factor(class), 
           linetype=as.factor(class))) +
  geom_density() +
  geom_vline(xintercept = 0.45, colour = "purple",
             linetype = 2)
  ggtitle("PC1 Scores by Class (Benign code=2 vs Malignant code=4)")

# look at confusion matrix
knitr::kable(table(bcdat$class,bcdat$pc1malig),
             caption = "Predicted Cases as Malignant (code=4) Based on PC1 Scores")
```

So, only 18 cases were miscoded out of 683, which is only 2.6% misclassified - not bad for an unsupervised PCA!

-----

6. Please save your RMD to a Github repository. Submit the PDF report for Homework 8 to CANVAS and include a link to your Homework 8 Github Repository.

**ANSWER KEY** This Homework Assignment and Associated Answer Key are located in the Github Repository at [https://github.com/melindahiggins2000/N741hmwk8](https://github.com/melindahiggins2000/N741hmwk8).

