---
title: "N741 Homework 8"
author: "Craig Alder"
date: "April 12, 2017"
output:
  github_document: default
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Homework 8 - DUE April 12, 2017 at 5pm

Please submit Homework 8 as a PDF to CANVAS no later than 5pm EST on April 12, 2017.

## Wisconsin Breast Cancer Data (Original)

For Homework 8 you will be working with the "Original" Wisconsin Breast Cancer dataset from the UCI Machine Learning Repository; see UCI dataset [http://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Original%29](http://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Original%29).

The raw data files can be downloaded from the associated Data Folder at [http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/](http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/). In this homework you will be working with the "breast-cancer-wisconsin.data" dataset, which is a CSV comma delimited file with NO column names in the 1st row. The datafile description and associated column file names are in the "breast-cancer-wisconsin.names" which is a simple text file [http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names](http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names). In this text file, as you read through it and scroll down, you'll see the following:

```
7. Attribute Information: (class attribute has been moved to last column)

   #  Attribute                     Domain
   -- -----------------------------------------
   1. Sample code number            id number
   2. Clump Thickness               1 - 10
   3. Uniformity of Cell Size       1 - 10
   4. Uniformity of Cell Shape      1 - 10
   5. Marginal Adhesion             1 - 10
   6. Single Epithelial Cell Size   1 - 10
   7. Bare Nuclei                   1 - 10
   8. Bland Chromatin               1 - 10
   9. Normal Nucleoli               1 - 10
  10. Mitoses                       1 - 10
  11. Class:                        (2 for benign, 4 for malignant)
```

So, the final datafile will have 11 columns. The dataset itself is a compilation of multiple groups of clinical cases also detailed in the breast-cancer-wisconsin.names" file [http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names](http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names).

The combined dataset has 699 cases (rows). However, 16 cases were missing values for the "Bare Nuclei" measurement. The R code below, processes the data, applies the names, and removes the cases with missing values. So, the final dataset created below `bcdat` will have 683 cases and 11 variables.

```{r}
# from tidyverse - use readr
# to read in the comma delimited dataset
library(readr)

# raw data does not have column names
bcdat <- read_csv("breast-cancer-wisconsin.data",
                  col_names=FALSE)

# add variable names
names(bcdat) <- c("idnum","clumpthickness","uniformcellsize",
                  "uniformcellshape","marginaladhesion",
                  "singlecellsize","barenuclei","blandchromatin",
                  "normalnucleoli","mitoses","class")

# note in column 7 "Bare Nucleoli" there are
# question marks "?" that need to be set to missing NA
library(dplyr)
bcdat <- bcdat %>%
  mutate(barenucfix = ifelse(barenuclei=="?",NA,
                              as.numeric(barenuclei)))

# keep the main 11 variables
bcdat <- bcdat %>%
  select(idnum,clumpthickness,uniformcellsize,uniformcellshape,
         marginaladhesion,singlecellsize,barenucfix,blandchromatin,  
         normalnucleoli,mitoses,class)

# keep only complete cases, n=683
bcdat <- na.omit(bcdat)
```

## Principal Components Analysis

For this Homework, please refer back to the code and exercises that Dr. Hertzberg presented during lesson 10 - specifically review towards the end of "Lesson10Part3.Rmd" see [https://github.com/vhertzb/Lesson10/blob/master/Lesson10Part3.Rmd](https://github.com/vhertzb/Lesson10/blob/master/Lesson10Part3.Rmd). During this exercise, Dr. Hertzberg introduced you to the `prcomp` procedure for performing principal components analysis. `prcomp` is part of the built-in `stats` package with base R. To learn morn type `help(prcomp)`.

In Dr. Hertzberg's example, she provided code for:

* performing the principal components analysis (pca)
* using the pca output to make a plot of the variances for each principal component (pc)
* computing the PVE (percent variance explained) and plotting the PVE
* and plotting the principal component "scores" of the cases (e.g. the "scores" plot)

I will layout the code below for running the PCA for the dataset as a whole, which will include also making a "loadings" plot for the variable "coefficients" or "loading weights" for each PC - these "loading plots" give us additional insight into (a) how the variables cluster or relate/correlate with each other or not and (b) where they fall in terms of relevance for each PC in the plot. For this dataset, we can easily get away with keeping only 2 PCs and making simplier 2D scatterplots for both the "loading plot" and "scores plot".

Use the code steps below to help you complete this homework 8 assignment.

## 1. Perform the PCA

```{r}
# use only columns 2 through 10
# you do not need the idnum, nor the class variables
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Make plots of the variance and PVE 

### Plot of the Variances of Each PC

```{r}
plot(pr.out)
```

### Plot of the PVE and Cumulative PVE of each PC

```{r}
# plots of the PVE percent variance explained
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Make a "loadings plot" of the variables

```{r}
# loadings are in the "rotation" part of the 
# pr.out list object. "rotation" is a matrix
# with a row for each variable and a column for
# each PC.
pr.out$rotation

# choose the 1st and 2nd columns for the 1st 2 PCs
# and plot these loading weights for the 9
# variables. I tweaked the limits some
# feel free to change these as needed
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")

# add xpd=FALSE to prevent lines drawn outside plot area
par(xpd=FALSE)

# add red dashed lines for the axes at y=0 and x=0
abline(h=0, col="red")
abline(v=0, col="red")

# overlay the variable names on this loading plot
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Scores Plot on 1st 2 PCs

```{r}
# scores plot - use x from the pr.out list object
# plot scores on 1st 2 PCs, columns 1 and 2 of x
# color the points by the "class" variable for
# benign (class=2) or malignant (class=4)
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdat$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2",
     sub = "Blue=Benign (class=2) and Red=Malignant (class=4)")
```

## Homework 8 Tasks

1. Rerun the PCA (steps 1-4 above) for (A) just the Benign cases and for just the (B) Malignant Cases. The code below, sets up these data subsets for you.

```{r}
# Benign cases
bcdatBenign <- bcdat %>%
  filter(class == 2)

# Malignant cases
bcdatMalignant <- bcdat %>%
  filter(class == 4)
```

_HINT: simply rename the new subsets and run the code steps above._ 

```{r}
# redo for benign ==============
bcdat <- bcdatBenign
# run steps above

# redo for malignant ==================
bcdat <- bcdatMalignant
```

###############################################

## 1. Benign, PCA

```{r}
pr.out <- prcomp(bcdatBenign[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Benign, plots of the variance and PVE 

### Plot of the Variances of Each PC

```{r}
plot(pr.out)
```

### Benign, plot of the PVE and Cumulative PVE of each PC

```{r}
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Benign, "loadings plot" of the variables

```{r}
pr.out$rotation
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")
par(xpd=FALSE)
abline(h=0, col="red")
abline(v=0, col="red")
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Benign, plot on 1st 2 PCs

```{r}
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdatBenign$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2")
```

## 1. Malignant, PCA

```{r}
pr.out <- prcomp(bcdatMalignant[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Malignant, plots of the variance and PVE 

### Plot of the Variances of Each PC

```{r}
plot(pr.out)
```

### Malignant, plot of the PVE and Cumulative PVE of each PC

```{r}
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Malignant, "loadings plot" of the variables

```{r}
pr.out$rotation
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")
par(xpd=FALSE)
abline(h=0, col="red")
abline(v=0, col="red")
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Malignant, plot on 1st 2 PCs

```{r}
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdatMalignant$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2")
```

2. In the overall dataset, when looking at the loadings plot, which variables cluster together? which variables do not lie with that cluster?

In PC1, the weights are all similar, indicating they all hang together. The mitoses variable has the lowest weight, but it does not seem too extreme.

In PC2 Mitoses takes up most of the component weight at .91. Barenucfix and blandchromatin are closest with weights of -.26 and -.23 respectively. Uniformcellsize, uniformcellshape, and normalnucleoli all have very low weights. 

In PC3 Clumpthickness takes up most of the weight at .86. Marginaladhesion is also substantive with a weight of .41. Uniformcellsize and uniformcellshape, and barenucfix all have extremely low weights.

These first three factors seem to show the general split. 

3. How do the variable clusters seen in the loading plots for the Benign data subset and Malignant subset differ? and how are they similar if at all?

Benign

-The first factor shows that, unlike the complete dataset, not all variables hang together. Mitoses has a very low weight compared to the other variables at -.04 (most hang out between -.35 and -.45). Clumpthickness and blandchromatin are also lower than the rest, though not too far off at .24 for each.
-The second factor is similar to the second for the entire dataset in that mitoses is by far the most heavily weighted at .83. Also, blandchromatin and barenucfix are weighted the next highest at .27 and .31 respectively, just as in the entire dataset.
-Blandchromatin and, unlike in the entire dataset, clumpthickness have the highest weights in PC3 at .6 and .62 respectively. 

Malignant

-The first factor shows that, unlike the complete dataset, not all variables hang together. Clumpthickness and barenucfix have very low weights compared to the other variables at -.04 and -.03 respectively. Clumpthickness was a little bit lower, but not extremely lower, in the benign loadings. 
-The second factor is different than both the benign and combined datasets. Barenucfix has the highest weight at -.57 while clumpthickness, marginal adhesion, blandchromatin, and mitoses all have substantial loadings (mitosis the smallest of this group at .29). Uniformcellsize, uniformcellshape, and normalnucleoli all have very small weights.
-Clumpthickness has a high weight for PC3 at -.72. Barenucfix, normalnucleoli, and mitosis also have substantial weights.

4. Is using 2 principal components reasonable for summarizing the variability seen in this Breast Cancer dataset with 9 measurements? Explain your reasoning for (a) the overall dataset, (b) the Benign subset and (c) the Malignant subset

2 principal components is reasonable for the overall dataset and the malignant subset, but three seems more reasonable for the benign subset.

Malignant subset: Each variable has a weight above .3 in one of the two components. Mitoses is the one exception with weights of -.27 in the first and .29 in the second, which seems reasonable for the second.

Benign subset: Each of the variabls has a weight of .3 or above in the first two components except blandchromatin and clump thickness, which max out at -.25 and -.27 respectively. In PC 3, they both have weights above .6.

Full dataset: Each variable has a weight of .3 or higher in one of the first two components.

5. While PCA is an unsupervised data analysis method (i.e. no "target" class information is used in the analysis), do you think the 2 PCs extracted do a good job of helping to distinguish Benign cases from Malignant cases (i.e. look back at the overall dataset Scores Plot). Explain your rationale.

There is very clearly a dense grouping of malignant cases in the plot, with very little benign cases mixed in, indicating the two PCs do a good job distinguishing cases.

6. Please save your RMD to a Github repository. Submit the PDF report for Homework 8 to CANVAS and include a link to your Homework 8 Github Repository.

Github repository: [https://github.com/craigalder/N741hmwk8](https://github.com/craigalder/N741hmwk8)
